---
title: "BTSBM: Bayesian Bradley-Terry Stochastic Block Models in R"
author: "Lapo Santi and Nial Friel"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  rmarkdown::pdf_document:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{BTSBM: A Software Paper}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 6.5,
  fig.height = 5,
  out.width = "88%",
  warning = FALSE,
  message = FALSE
)
set.seed(123)
```

# Abstract {-}

This software paper documents BTSBM, an R package for Bayesian Bradley-Terry stochastic block models in paired comparison problems. The package is built for settings where strict item-by-item ranking is often too brittle, but an ordered block structure is stable and interpretable. We model directed wins \(w_{ij}\), infer latent allocations \(x_i\), infer the number of occupied blocks \(K\), and estimate block intensities \(\lambda_k\). The package provides Gibbs samplers, post-processing for label alignment, visual diagnostics, and model comparison tools based on leave-one-out criteria.

The paper has two aims. The first is methodological: to explain the model and the Markov chain Monte Carlo updates in the notation used across the package. The second is practical: to provide a reproducible workflow with code and output, in the style commonly used in R package vignettes.

```{r lib}
library(BTSBM)
```

# Introduction

Paired comparison data appear in sports, psychometrics, preference studies, and many ranking tasks. In these problems, we observe directed outcomes such as wins of item \(i\) over item \(j\), stored as \(w_{ij}\). A standard Bradley-Terry formulation associates each item with a positive latent intensity and converts intensity contrasts into win probabilities.

The strict item-level view can be too fine. Two players may be practically indistinguishable in ability, even if the model can force one above the other. The same pattern appears in many ranking contexts. A block model addresses this by grouping items into latent clusters and ordering these clusters by intensity. In BTSBM this is done in a Bayesian way, so uncertainty on allocations and on ranking is explicit.

# Model and Notation

We follow the notation used in the package vignette on symbols and dimensions.

Let \(w_{ij}\) be directed wins, with \(w_{ii}=0\), and define
\[
 n_{ij}=w_{ij}+w_{ji}.
\]

Each item \(i\) has a latent block label \(x_i \in \{1,\dots,K\}\). Each occupied block \(k\) has intensity \(\lambda_k>0\). The Bradley-Terry map is
\[
\theta_{x_i x_j} = \frac{\lambda_{x_i}}{\lambda_{x_i}+\lambda_{x_j}},
\]
and outcomes satisfy
\[
w_{ij} \mid n_{ij}, x, \lambda \sim \mathrm{Binomial}(n_{ij}, \theta_{x_i x_j}).
\]

For priors on \(\lambda_k\), BTSBM uses
\[
\lambda_k \sim \mathrm{Gamma}(a,b), \qquad b = \exp\{\psi(a)\},
\]
so that \(\mathbb{E}[\log \lambda_k]=0\) a priori.

For partitions, the package supports DP, PY, DM, and GN urn-style priors through prior weights in the single-site allocation updates.

## Priors on \(x\): what is supported and how to set it

The prior on allocations \(x=(x_1,\ldots,x_n)\) is selected with the `prior` argument in `gibbs_bt_sbm()`. The package currently supports four options, aligned with the exchangeable partition constructions used in Legramanti, Rigon, Durante, and Dunson (2022):

1. `prior = "DP"` (Dirichlet process / CRP-type weights)
  Uses `alpha_PY` and ignores `sigma_PY`:
  \[
  (v_1,\ldots,v_K,\alpha_{PY}),
  \]
  implemented by `urn_DP(v_minus, alpha_PY)`.

2. `prior = "PY"` (Pitman-Yor)
  Uses `alpha_PY` and `sigma_PY \in (0,1)`:
  \[
  (v_1-\sigma_{PY},\ldots,v_K-\sigma_{PY},\alpha_{PY}+K\sigma_{PY}),
  \]
  implemented by `urn_PY(v_minus, alpha_PY, sigma_PY)`.

3. `prior = "DM"` (finite Dirichlet-Multinomial)
  Uses `beta_DM` and a finite cap `K_DM`:
  \[
  (v_1+\beta_{DM},\ldots,v_K+\beta_{DM},\beta_{DM}(K_{DM}-K)\mathbf{1}_{K<K_{DM}}),
  \]
  implemented by `urn_DM(v_minus, beta_DM, K_DM)`.

4. `prior = "GN"` (Gnedin-type finite-species prior)
  Uses `gamma_GN`:
  \[
  \big((v_1+1)(n-K+\gamma),\ldots,(v_K+1)(n-K+\gamma),K^2-K\gamma\big),
  \]
  implemented by `urn_GN(v_minus, gamma_GN)`.

In code, `gibbs_bt_sbm()` validates the needed hyperparameters and then dispatches through
`urn_fun <- switch(prior, DP = ..., PY = ..., DM = ..., GN = ...)`.
During each single-site update of \(x_i\), this gives the prior mass for assigning item \(i\) to each existing block and to one candidate new block.

Minimal usage examples:

```{r prior-examples, eval=FALSE}
# Dirichlet process prior on x
fit_dp <- gibbs_bt_sbm(w_ij, prior = "DP", alpha_PY = 1.0)

# Pitman-Yor prior on x
fit_py <- gibbs_bt_sbm(w_ij, prior = "PY", alpha_PY = 1.0, sigma_PY = 0.3)

# Finite Dirichlet-Multinomial prior on x (at most K_DM blocks)
fit_dm <- gibbs_bt_sbm(w_ij, prior = "DM", beta_DM = 0.5, K_DM = 15)

# Gnedin prior on x
fit_gn <- gibbs_bt_sbm(w_ij, prior = "GN", gamma_GN = 0.8)
```

These four choices correspond directly to the helpers `urn_DP`, `urn_PY`, `urn_DM`, and `urn_GN` used internally by the sampler, and are the package-level interface to the partition-prior family discussed in Legramanti, Rigon, Durante, and Dunson (2022).

# Posterior Computation

The core sampler in `gibbs_bt_sbm()` alternates three main blocks.

First, latent variables \(Z_{ij}\) are sampled for observed pairs \((i,j)\) with \(n_{ij}>0\):
\[
Z_{ij} \sim \mathrm{Gamma}\!\left(n_{ij}, \frac{1}{\lambda_{x_i}+\lambda_{x_j}}\right),
\]
with symmetry \(Z_{ij}=Z_{ji}\). This step is implemented in C++ for speed.

Second, each label \(x_i\) is updated by single-site Gibbs moves. Existing clusters and one new-cluster option are compared on a log-weight scale:
\[
\log p_h \propto \log(\text{prior mass}_h) + \text{likelihood contribution}_h.
\]
For existing clusters, the key term is
\[
w_i\log\lambda_h - \lambda_h\sum_j Z_{ij}, \qquad w_i = \sum_j w_{ij}.
\]
For the new cluster, the package uses the integrated score under the Gamma prior.

Third, occupied \(\lambda_k\) are updated conjugately:
\[
\lambda_k \mid - \sim \mathrm{Gamma}\left(a + \sum_{i:x_i=k} w_i,\ b + \sum_{i:x_i=k}\sum_j Z_{ij}\right).
\]
Then a global geometric-mean rescaling of occupied intensities is applied for numerical stability.

# Software Design

The implementation separates fitting, utilities, and plotting.

- `gibbs_bt_sbm()` is the clustered sampler.
- `gibbs_bt_simple()` is the item-level Bradley-Terry baseline.
- `relabel_by_lambda()` aligns labels across MCMC draws and computes posterior partition summaries.
- `plot_block_adjacency()`, `plot_assignment_probabilities()`, `plot_lambda_uncertainty()`, and `plot_rank_intervals()` provide structured diagnostics.
- `make_bt_simple_loo()`, `make_bt_cluster_loo()`, and `compare_bt_models_loo()` provide out-of-sample comparison helpers.

The package ships with `ATP_2000_2025`, a yearly list with `Y_ij`, `N_ij`, and player metadata.

# Reproducible Workflow

## Simulation-first example

To ensure clear multi-cluster structure in a short run, we begin with synthetic paired-comparison data generated from three latent blocks.

```{r sim-data}
sim <- sample_from_BTSBM(
  n_players = 18,
  K = 3,
  mean_matches = 8,
  p_edge = 0.75,
  lambda = c(0.25, 0.9, 2.4),
  reverse_lambda = FALSE,
  seed = 123
)

w_ij <- sim$w
diag(w_ij) <- 0L
player_ids <- paste0("Player_", seq_len(nrow(w_ij)))
dimnames(w_ij) <- list(player_ids, player_ids)

w_ij[1:6, 1:6]
```

```{r dims-sim}
n <- nrow(w_ij)
S_demo <- 300 - 150
c(n = n, S = S_demo)
```

## Fit clustered model

```{r fit-cluster}
fit <- gibbs_bt_sbm(
  w_ij = w_ij,
  a = 2,
  prior = "GN",
  gamma_GN = 0.8,
  T_iter = 300,
  T_burn = 150,
  verbose = FALSE
)

str(list(
  x_samples = dim(fit$x_samples),
  K_per_iter = length(fit$K_per_iter),
  lambda_samples = length(fit$lambda_samples)
))
```

```{r K-quick-check}
head(sort(table(fit$K_per_iter), decreasing = TRUE), 6)
```

## Relabel and summarize

```{r relabel}
post <- relabel_by_lambda(fit$x_samples, fit$lambda_samples)
if (knitr::is_latex_output()) {
  knitr::kable(
    post$block_count_distribution,
    caption = "Posterior distribution of the number of occupied blocks"
  )
} else {
  pretty_table_K_distribution(post)
}
```

```{r partitions}
head(post$block_count_distribution)
```

```{r cluster-count}
length(unique(post$minVI_partition))
```

## Visual diagnostics

```{r plot-block, fig.cap="Block-ordered adjacency heatmap. Rows and columns are ordered by inferred blocks, with boundary lines indicating block transitions."}
plot_block_adjacency(fit = post, w_ij = w_ij)
```

```{r plot-assign, fig.height=6, fig.cap="Posterior assignment probabilities by player and cluster, after relabelling for interpretability."}
plot_assignment_probabilities(fit = post, w_ij = w_ij, x_hat = post$minVI_partition)
```

```{r plot-lambda, fig.height=6, fig.cap="Posterior uncertainty in block intensity parameters (lambda), summarising relative block strengths."}
plot_lambda_uncertainty(fit = post, w_ij = w_ij, x_hat = post$minVI_partition)
```

```{r plot-rank, fig.height=6, fig.cap="Posterior rank intervals for the top players, highlighting uncertainty in the induced ranking."}
plot_rank_intervals(post, max_players = 10)
```

## Small ATP illustration

The simulation above is used for cluster-recovery diagnostics. For a lightweight real-data illustration, we run the same workflow on a small ATP subset.

```{r atp-data}
w_ij_atp <- ATP_2000_2025$`2017`$Y_ij[1:10, 1:10]
diag(w_ij_atp) <- 0L
atp_ids <- rownames(w_ij_atp)
if (is.null(atp_ids)) {
  atp_ids <- paste0("ATP_", seq_len(nrow(w_ij_atp)))
}
dimnames(w_ij_atp) <- list(atp_ids, atp_ids)

fit_atp <- gibbs_bt_sbm(
  w_ij = w_ij_atp,
  a = 2,
  prior = "GN",
  gamma_GN = 0.8,
  T_iter = 180,
  T_burn = 90,
  verbose = FALSE
)

post_atp <- relabel_by_lambda(fit_atp$x_samples, fit_atp$lambda_samples)
head(post_atp$block_count_distribution)
```

## Predictive comparison with simple BT

```{r fit-simple}
fit_simple <- gibbs_bt_simple(
  w_ij = w_ij,
  a = 0.01,
  b = 0.1,
  T_iter = 300,
  T_burn = 150,
  verbose = FALSE
)

llo_simple <- make_bt_simple_loo(w_ij, fit_simple$lambda_samples)
llo_cluster <- make_bt_cluster_loo(
  w_ij,
  post$lambda_samples_relabel,
  post$x_samples_relabel
)

# The compare step requires the suggested package 'loo'.
if (requireNamespace("loo", quietly = TRUE)) {
  cmp_out <- tryCatch(
    compare_bt_models_loo(llo_simple, llo_cluster),
    error = function(e) e
  )

  if (inherits(cmp_out, "error")) {
    paste(
      "LOO comparison is unavailable in this local setup:",
      conditionMessage(cmp_out)
    )
  } else {
    cmp_out$comparison
  }
} else {
  "Install 'loo' to run compare_bt_models_loo()."
}
```

# Practical Notes for Package Use

The object dimensions used by BTSBM are consistent across functions. For \(S=T_{\text{iter}}-T_{\text{burn}}\), `x_samples` has shape \(S \times n\), `K_per_iter` has length \(S\), and relabelled item-level intensities are stored as `lambda_samples_relabel` with shape \(S \times n\).

The main plotting functions expect relabelled output, not raw labels, because raw cluster indices are permutation-invariant across iterations.

For computational control, the most important switches are matrix size \(n\), number of saved draws \(S\), and whether latent \(Z\) draws are stored. The `store_z = TRUE` option is useful for diagnostics but substantially increases memory use.

# Assumptions and Limitations

BTSBM is intentionally focused. Outcomes are modelled through Bradley-Terry probabilities determined by latent intensities; pair-specific covariates and dynamic effects are not in the current core model. Partition priors are exchangeable and encoded through urn-style updates. The approach is effective for moderate problem sizes common in sports and ranking applications, but very large dense comparison matrices may require additional engineering.

# Conclusion

BTSBM provides a coherent workflow for Bayesian block-structured ranking from paired comparisons. It combines a model with interpretable latent structure, practical posterior computation, and publication-oriented diagnostics. The package is especially useful when strict total rankings are unstable but tiered structure is statistically and substantively credible.

# References

Bradley, R. A. and Terry, M. E. (1952). Rank analysis of incomplete block designs: I. The method of paired comparisons. Biometrika, 39(3/4), 324-345.

Ferguson, T. S. (1973). A Bayesian analysis of some nonparametric problems. The Annals of Statistics, 1(2), 209-230.

Pitman, J. and Yor, M. (1997). The two-parameter Poisson-Dirichlet distribution derived from a stable subordinator. The Annals of Probability, 25(2), 855-900.

Legramanti, S., Rigon, T., Durante, D., and Dunson, D. B. (2022). Extended stochastic block models with application to criminal networks. The Annals of Applied Statistics, 16(2), 865-891.

Vehtari, A., Gelman, A., and Gabry, J. (2017). Practical Bayesian model evaluation using leave-one-out cross-validation and WAIC. Statistics and Computing, 27, 1413-1432.
