---
title: "Getting started: from results to a Bradley--Terry analysis"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Getting started: from results to a Bradley--Terry analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

## What this package does

BTSBM analyses two kinds of preference data:

1. Pairwise results: one item wins or loses against another, such as a match
   in a league.
2. Rankings: a person places several items from most to least preferred.

This first tutorial uses pairwise results. The ranking tutorial is
vignette("pl-ranking-models").

A Bradley--Terry (BT) model assigns every player a positive strength. A larger
strength means a greater probability of winning against a player with a smaller
strength. A BT--SBM additionally lets the data group players with similar
strengths into latent tiers.

```{r library}
library(BTSBM)
```

## 1. Start with data you can inspect

The toy tennis league is synthetic and intentionally small. wins[i, j] means
the number of times player i beat player j. The diagonal is zero because a
player does not play themself.

```{r make-data}
tennis <- tennis_toy(matches_per_pair = 8, seed = 1)
tennis$wins
```

Before fitting a model, inspect the observed outcomes. Darker cells identify
frequent wins by the row player over the column player.

```{r observed-outcomes, fig.height=5}
plot_pairwise_outcomes(tennis)
```

This plot is descriptive: it does not yet account for uncertainty or combine
all opponents into one estimate of player strength.

## 2. Choose a model

Start with a BT--SBM when you think players may form broad tiers rather than a
completely reliable order from first to last. Here, the Gnedin prior lets the
number of tiers be learned from the data.

```{r choose-model}
model <- bt_sbm_model(
  latent_strength = latent_strength(shape = 4, rate = exp(digamma(4))),
  item_clustering = gnedin_prior(gnedin_hyperparameter = 0.5)
)
model
```

The latent_strength() prior must be positive. item_clustering is a prior over
possible player tiers. You can use finite_partition(max_clusters = 3) when a
strict upper limit is scientifically justified.

## 3. Fit the model

An MCMC fit produces many plausible values rather than one single answer.
warmup draws are discarded; the remaining draws represent posterior
uncertainty. The short chain below is suitable for a tutorial. Use longer,
multiple chains in a real study.

```{r fit}
fit <- fit_btsbm(
  tennis,
  model,
  mcmc_control(iter = 400, warmup = 200, seed = 2)
)
fit
```

## 4. Read the main result: strengths and uncertainty

The point is the posterior mean strength. The horizontal line is a 90% credible
interval: it describes the remaining uncertainty after seeing the results.
Intervals that overlap substantially should not be read as a certain ordering.

```{r strength-plot, fig.height=4.5}
plot_strength_summary(fit)
```

The same information is available as a table, which is useful for reporting.

```{r strength-table}
strength_summary(fit)
```

## 5. Ask whether the data support tiers

The model's cluster labels can switch their names from one MCMC draw to the
next. Instead, use a posterior similarity plot. A dark cell means the two
players were frequently placed in the same tier.

```{r similarity-plot, fig.height=5}
plot_posterior_similarity(fit, target = "item")
```

## 6. Check the sampler before interpreting it

MCMC draws should move around a stable region. Trace plots are a quick
screening tool: long trends or a completely stuck trace are reasons to run
longer chains or reconsider the model.

```{r trace-plot, fig.height=4}
plot_mcmc_traces(fit)
mcmc_diagnostics(fit)
```

## Where to go next

* Use bt_model() if every item should have its own strength without tiers.
* Use vignette("pl-ranking-models") when the observations are full or top-m
  rankings rather than wins and losses.
* Use vignette("toy-data-gallery") for additional worked synthetic examples.
