---
title: "Ranking data: a beginner's guide to Plackett--Luce models"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Ranking data: a beginner's guide to Plackett--Luce models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

## When should I use a ranking model?

Use a Plackett--Luce (PL) model when each observation is an ordered list, such
as sushi A was first, sushi B was second, and sushi C was third. This is
different from pairwise data, where an observation only says who won one
head-to-head comparison.

The model gives every item a positive relative strength. At each position in a
ranking, items with larger strengths are more likely to be selected. The
strengths are relative rather than absolute: multiplying every strength by the
same number would not change the predicted rankings.

```{r library}
library(BTSBM)
```

## 1. Examine rankings before fitting

The original Sushi Preference Dataset is a well-known benchmark. Its licence
does not allow redistribution, so BTSBM includes a small synthetic
Sushi-inspired teaching dataset instead. You can obtain the original data from
[Kamishima's data page](https://www.kamishima.net/soft/) if its terms suit your
use.

```{r sushi-data}
sushi <- sushi_toy(n_rankings = 40, rank_length = 5, seed = 2026)
sushi
sushi$rankings[1:4, ]
```

Each row is one respondent and each column is a reported position. The next
plot is a useful first look. It counts how often an item appears in each
position; it is not yet a fitted model.

```{r ranking-positions}
plot_ranking_positions(sushi)
```

## 2. Fit a simple PL model

Start with the simplest model: one strength for each sushi type. The Gamma
prior keeps strengths positive. The short chain below is for illustration; use
longer, multiple chains in a real analysis.

```{r pl-fit}
simple_pl <- fit_btsbm(
  sushi,
  pl_model(latent_strength = latent_strength(shape = 1, rate = 1)),
  mcmc_control(iter = 600, warmup = 300, seed = 1)
)
simple_pl
```

This plot is usually the main analysis result. The point is the posterior mean
relative strength; the interval displays posterior uncertainty.

```{r simple-pl-strengths, fig.height=5}
plot_strength_summary(simple_pl)
```

Rank intervals translate those strengths into an easier question: what place
could each sushi item plausibly occupy? Rank 1 is best.

```{r simple-pl-ranks, fig.height=5}
plot_rank_intervals(simple_pl)
```

## 3. Do some sushi types form a tier?

A PL--SBM lets items share a latent strength. It is useful when the data support
broad groups, such as often favoured and rarely favoured, but do not support a
precise order inside every group.

```{r pl-sbm-fit}
tiered_pl <- fit_btsbm(
  sushi,
  pl_sbm_model(
    item_clustering = gnedin_prior(gnedin_hyperparameter = 0.5)
  ),
  mcmc_control(iter = 600, warmup = 300, seed = 2)
)
```

Do not read raw MCMC cluster labels: label 1 in one draw need not mean the same
thing as label 1 in another draw. Instead, interpret the posterior similarity
plot. A dark cell means the two sushi types were often placed in the same tier.

```{r item-similarity, fig.height=6}
plot_posterior_similarity(tiered_pl, target = "item")
```

You can choose other priors for the item grouping. The names describe the
decision being made:

```{r priors, eval=FALSE}
dirichlet_process_prior(concentration = 1)
pitman_yor_prior(concentration = 1, discount = 0.2)
finite_partition(max_clusters = 4, concentration = 1)
```

## 4. Could respondents have different preference profiles?

A PL mixture groups rankings, not sushi items. It answers questions such as
whether there are distinct kinds of respondent preference. The
Eurovision-inspired toy data is especially useful here because it has simulated
jury and viewer profiles.

```{r mixture-data}
votes <- eurovision_toy(n_rankings = 36, rank_length = 5, seed = 3)
plot_ranking_positions(votes)
```

```{r pl-mixture-fit}
mixture_fit <- fit_btsbm(
  votes,
  pl_mixture_model(
    ranking_clustering = dirichlet_process_prior(concentration = 1)
  ),
  mcmc_control(iter = 400, warmup = 200, seed = 4)
)
```

The rows and columns now refer to rankings, not acts. Dark blocks suggest
respondents whose preferences were often assigned together.

```{r ranking-similarity, fig.width=7, fig.height=7}
plot_posterior_similarity(mixture_fit, target = "ranking")
```

## 5. Jointly group respondents and items

The PL--LBM has both kinds of grouping: it clusters ranking rows and items at
the same time. It is useful for a question like which respondent profiles
prefer which groups of items. pl_lbm_toy() has known simulated structure so
you can see the workflow.

```{r pl-lbm-fit}
structured_rankings <- pl_lbm_toy(n_rankings = 24, rank_length = 4, seed = 5)
lbm_fit <- fit_btsbm(
  structured_rankings,
  pl_lbm_model(
    ranking_clustering = finite_partition(max_clusters = 3),
    item_clustering = finite_partition(max_clusters = 3)
  ),
  mcmc_control(iter = 400, warmup = 200, seed = 6)
)
```

```{r lbm-item-similarity, fig.height=5}
plot_posterior_similarity(lbm_fit, target = "item")
```

```{r lbm-ranking-similarity, fig.width=7, fig.height=7}
plot_posterior_similarity(lbm_fit, target = "ranking")
```

## 6. Check the MCMC run

The diagnostics table and traces should be checked before interpreting a
partition. They are not a substitute for multiple chains, but they make it
easy to spot a trace that is stuck or still drifting.

```{r diagnostics, fig.height=5}
mcmc_diagnostics(tiered_pl)
plot_mcmc_traces(tiered_pl)
```

For model comparison, log_lik() returns one likelihood contribution per
ranking. If the optional loo package is installed, use loo_btsbm() to perform
ranking-level PSIS-LOO.

## References

* Caron, F. and Doucet, A. (2012). Efficient Bayesian inference for
  generalized Bradley--Terry models. *Journal of Computational and Graphical
  Statistics*, 21(1), 174--196. <https://doi.org/10.1080/10618600.2012.638220>
* Kamishima, T. (2003). Nantonac collaborative filtering: Recommendation based
  on order responses. *Proceedings of the Ninth ACM SIGKDD International
  Conference on Knowledge Discovery and Data Mining*.
