---
title: "Plackett--Luce ranking models"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Plackett--Luce ranking models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Ranking data

The Plackett--Luce (PL) likelihood models a strict ranking as a sequence of
choices from a shrinking set of items.  `BTSBM` accepts the compact `rho`
format used by the PLuce reference implementation: each row is one ranking and
columns run from most to least preferred item.  A row may be a full ranking or
a top-`m` prefix, provided every ranking uses the same item universe.

The initial sampler uses the exponential-race augmentation of Caron and Doucet
(2012).  It yields Gamma conditional updates for positive abilities while
preserving the PL likelihood.  This is also the computational basis for the
item-block PL--SBM implementation.

## A Sushi-inspired demonstration

The original Sushi Preference Dataset is a well-known ranking benchmark: it
contains surveys of preferences over sushi and is described by Kamishima (2003).
The original licence does not allow redistribution, so this package does not
bundle any of its respondents' rankings.  `sushi_toy()` is instead a small,
reproducible synthetic teaching dataset with the same application setting and
two known preference profiles.  Obtain the original dataset directly from
[Kamishima's data page](https://www.kamishima.net/soft/) if its licence suits
your use.

```{r sushi-data}
library(BTSBM)

sushi <- sushi_toy(n_rankings = 40, rank_length = 5, seed = 2026)
sushi
sushi$rankings[1:4, ]
```

Fit a simple PL model with a Gamma prior on item latent strengths.
`posterior_strength()` returns reporting-scale strengths, normalised to have
geometric mean one by default.

```{r pl-fit}
fit_pl <- fit_btsbm(
  sushi,
  pl_model(latent_strength = latent_strength(shape = 1, rate = 1)),
  mcmc_control(iter = 600, warmup = 300, seed = 1)
)

posterior_strength(fit_pl, summary = "mean")
```

`log_lik()` has one column per ranking row, the conditionally independent unit
needed for ranking-level posterior predictive checks and PSIS-LOO.

```{r pl-loglik}
dim(log_lik(fit_pl))
```

## PL--SBM: item blocks with tied abilities

The PL--SBM shares a strength among items assigned to the same latent block.
The current implementation supports the Gnedin, Dirichlet process,
Pitman--Yor, and finite Dirichlet--multinomial item-partition priors through
named helpers. The names make the model intent explicit: `item_clustering`
describes which items share a strength, and `gnedin_hyperparameter` controls
the Gnedin prior.

```{r pl-sbm-fit}
fit_sbm <- fit_btsbm(
  sushi,
  pl_sbm_model(
    item_clustering = gnedin_prior(gnedin_hyperparameter = 0.5)
  ),
  mcmc_control(iter = 600, warmup = 300, seed = 2)
)

fit_sbm
fit_sbm$draws$n_item_clusters
posterior_similarity(fit_sbm, target = "item")
```

The block labels are exchangeable.  Use the posterior similarity matrix rather
than raw label frequencies when summarising the partition until a
model-specific canonicalisation method is chosen.

To choose another prior, substitute one of the following objects for
`item_clustering`:

```{r priors, eval=FALSE}
dirichlet_process_prior(concentration = 1)
pitman_yor_prior(concentration = 1, discount = 0.2)
finite_partition(max_clusters = 4, concentration = 1)
```

## Next PL models

`pl_mixture_model()` clusters ranking rows, while `pl_lbm_model()` jointly
clusters ranking rows and items with a `C x K` latent-strength matrix. Both
now have readable R Gibbs samplers based on the PLuce exponential-race
architecture. They are suitable for small and moderate teaching/data-analysis
problems; the PLuce C++ aggregation kernels are the next performance step.

```{r pl-mixture-fit}
mixture_fit <- fit_btsbm(
  sushi,
  pl_mixture_model(
    ranking_clustering = dirichlet_process_prior(concentration = 1)
  ),
  mcmc_control(iter = 300, warmup = 150, seed = 3)
)

posterior_similarity(mixture_fit, target = "ranking")
mcmc_diagnostics(mixture_fit)
```

The LBM is exercised with a known-structure synthetic fixture:

```{r pl-lbm-fit}
lbm_fit <- fit_btsbm(
  pl_lbm_toy(n_rankings = 20, rank_length = 4),
  pl_lbm_model(
    ranking_clustering = finite_partition(max_clusters = 3),
    item_clustering = finite_partition(max_clusters = 3)
  ),
  mcmc_control(iter = 300, warmup = 150, seed = 4)
)

mcmc_diagnostics(lbm_fit)
```

`mcmc_diagnostics()` reports effective sample-size estimates, Monte Carlo
standard errors, and lag-one autocorrelation for the total ranking likelihood
and available occupied-cluster traces. Use these alongside multiple chains and
posterior predictive checks for a substantive analysis.

`pl_lbm_toy()` provides a structural fixture with known ranking and item
partitions for diagnostics and forthcoming PLuce parity tests.

## References

* Caron, F. and Doucet, A. (2012). Efficient Bayesian inference for
  generalized Bradley--Terry models. *Journal of Computational and Graphical
  Statistics*, 21(1), 174--196. <https://doi.org/10.1080/10618600.2012.638220>
* Kamishima, T. (2003). Nantonac collaborative filtering: Recommendation based
  on order responses. *Proceedings of the Ninth ACM SIGKDD International
  Conference on Knowledge Discovery and Data Mining*.
