---
title: "Toy datasets: choose a model and read its output"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Toy datasets: choose a model and read its output}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

This gallery is a map of the package. Each dataset is synthetic, small enough
to run quickly, and designed around one practical question. They are for
learning and testing; they are not real contest, survey, or match data.

```{r library}
library(BTSBM)
```

## If you have win/loss counts: tennis

Use a Bradley--Terry model when every observation is a head-to-head outcome.
The first question is simply: which player beat which opponent most often?

```{r tennis-data}
tennis <- tennis_toy(matches_per_pair = 8, seed = 1)
plot_pairwise_outcomes(tennis)
```

Fit a BT--SBM if players may belong to broad tiers.

```{r tennis-fit}
tennis_fit <- fit_btsbm(
  tennis,
  bt_sbm_model(
    item_clustering = finite_partition(max_clusters = 3)
  ),
  mcmc_control(iter = 300, warmup = 150, seed = 2)
)
```

```{r tennis-output, fig.height=4.5}
plot_strength_summary(tennis_fit)
plot_posterior_similarity(tennis_fit, target = "item")
```

Read the first plot as relative player strength and uncertainty. Read the
second as evidence that pairs of players belong to the same tier.

## If you have rankings: Sushi-inspired preferences

Use a PL model when each row is an ordered list. First inspect which items
occur near the top of the reported rankings.

```{r sushi-data}
sushi <- sushi_toy(n_rankings = 30, rank_length = 5, seed = 3)
plot_ranking_positions(sushi)
```

Then estimate item strengths and their uncertainty.

```{r sushi-fit}
sushi_fit <- fit_btsbm(
  sushi,
  pl_model(),
  mcmc_control(iter = 300, warmup = 150, seed = 4)
)
```

```{r sushi-output, fig.height=5}
plot_strength_summary(sushi_fit)
plot_rank_intervals(sushi_fit)
```

The rank plot is particularly useful when strengths are close: it shows how
much the possible ordering changes across posterior draws.

## If respondents may have different tastes: Eurovision-inspired rankings

Here the items are fictional acts, and the ranking rows are simulated from two
different voting profiles. Use a PL mixture when the question is about groups
of ranking rows rather than groups of items.

```{r eurovision-data}
votes <- eurovision_toy(n_rankings = 30, rank_length = 5, seed = 5)
plot_ranking_positions(votes)
```

```{r eurovision-fit}
mixture_fit <- fit_btsbm(
  votes,
  pl_mixture_model(
    ranking_clustering = dirichlet_process_prior(concentration = 1)
  ),
  mcmc_control(iter = 300, warmup = 150, seed = 6)
)
```

```{r eurovision-output, fig.width=7, fig.height=7}
plot_posterior_similarity(mixture_fit, target = "ranking")
```

Rows and columns are individual rankings. Darker blocks are evidence of
rankings that repeatedly occur in the same latent preference profile.

## If both rows and items may have structure: PL--LBM

The PL latent block model simultaneously groups ranking rows and items. This is
the most structured option in the package. The synthetic fixture below has
known generating groups, so it is a safe place to learn the workflow.

```{r lbm-fit}
structured_rankings <- pl_lbm_toy(n_rankings = 20, rank_length = 4, seed = 7)
lbm_fit <- fit_btsbm(
  structured_rankings,
  pl_lbm_model(
    ranking_clustering = finite_partition(max_clusters = 3),
    item_clustering = finite_partition(max_clusters = 3)
  ),
  mcmc_control(iter = 300, warmup = 150, seed = 8)
)
```

```{r lbm-output, fig.height=5}
plot_posterior_similarity(lbm_fit, target = "item")
```

```{r lbm-traces, fig.height=5}
plot_mcmc_traces(lbm_fit)
mcmc_diagnostics(lbm_fit)
```

For every model, the final step is the same: inspect diagnostics, rerun with
longer and multiple chains for real data, and avoid treating a dark similarity
cell or a narrow interval as a guarantee.
